\section{Extended Analysis}
\label{app:extended}

While the results in Section~\ref{sec:comparison} are good,
it is desirable to have a more extensive analysis.
The only algorithms tested are
\UnrolledThree{} (Listing~\ref{list:newton-unrolled-3}),
\WhileOne{} (Listing~\ref{list:newton-while-1}),
\WhileTwo{} (Listing~\ref{list:newton-while-2}),
\WhileThree{} (Listing~\ref{list:newton-while-3}), and
\Linear{} (Listing~\ref{list:newton-linear}),
as these performed the best:
they had a significant number of values which were minimal.

The deterministic and random values will be separated.
All values tested in Section~\ref{sec:comparison}
are included in these tests.

\subsection{Deterministic Tests}
\label{app:deterministic}

\subsubsection{Overview}

We test additional deterministic values.
First, we include all deterministic values from Section~\ref{sec:comparison}.
Additional values are of the form $2^{k} + 2^{j}$ for $j,k\in\N$ and $j < k$.
The total number of values were 33154.
The results are shown in Table~\ref{table:minimal_gas_costs_ed}
and Figure~\ref{fig:minimal_gas_hist_ed}.

The minimal values have a somewhat different trend than before:
\UnrolledThree{}, \Linear{}, and \WhileOne{} all perform well.
By looking at Tables~\ref{table:minimal_gas_costs_ed_no_linear} and
\ref{table:minimal_gas_costs_ed_no_unrolled},
we see that \UnrolledThree{} and \Linear{} ``steal''
minimal values from each other.
A direct comparison between \UnrolledThree{} and \Linear{}
may be found in Table~\ref{table:minimal_gas_costs_ed_unrolled_linear},
where \UnrolledThree{} has the majority of minimum values.
The fact that \WhileOne{} performs the best in
Table~\ref{table:minimal_gas_costs_ed}
may be explained by the initialization value in this test
frequently starts \emph{very close} to the correct value;
see the discussion in Appendix~\ref{app:deterministic_detailed}.
It is interesting that \UnrolledThree{} and \Linear{} performed
as well as they did in this situation which favors \WhileOne{}.

The results here show that it is easy to (unwittingly)
test values which are biased toward one particular algorithm.
This is the reason for testing a large collection
of random values in Appendix~\ref{app:loguniform}.

\subsubsection{Detailed Look}
\label{app:deterministic_detailed}

We now take a closer look at particular values:
we look at all values $2^{254} \le v < 2^{255}$
and then $2^{255} \le v < 2^{256}$.
In this instance, we are just comparing the gas cost of \UnrolledThree{}
and \WhileOne{}.
We expect these two collections to be characteristic
of the remaining gas values.

\paragraph{Values $2^{254} \le v < 2^{255}$}
For \UnrolledThree{}, there are only two gas costs in this range:
$810$ and $817$.
\WhileOne{} starts off at a low gas cost of $572$;
it then jumps to $662$, $752$, $842$, $932$, and $1022$.
We estimate that each Newton iteration costs $90$ gas.

\paragraph{Values $2^{255} \le v < 2^{256}$}
For \UnrolledThree{}, the gas cost is predominantly $810$;
there is one instance of $817$, and three values of $168$ (from early exit).
Almost all gas cost values for \WhileOne{} are $1022$,
with some larger values.

\input{tables/minimal_gas_ed.tex}
\input{plots/minimal_hist_plots_ed.tex}

\subsection{Loguniform Tests}
\label{app:loguniform}

The tests in Section~\ref{sec:comparison} ran 1280 values
which came from a loguniform distribution~\cite{ScipyLoguniform}.
Here, we perform the same test using more samples;
in particular, we use $16384 = 2^{14}$ random values from 16808 samples.
The results are shown in Table~\ref{table:minimal_gas_costs_er}
and Figure~\ref{fig:minimal_gas_hist_er}.
In this case, both \UnrolledThree{} and \Linear{} perform well.
By looking at Tables~\ref{table:minimal_gas_costs_er_no_linear} and
\ref{table:minimal_gas_costs_er_no_unrolled},
we see that \UnrolledThree{} and \Linear{} ``steal''
minimal values from each other.
A direct comparison between \UnrolledThree{} and \Linear{}
may be found in Table~\ref{table:minimal_gas_costs_er_unrolled_linear},
where \UnrolledThree{} has the majority of minimum values.

There is no significant difference between the results in 
Tables~\ref{table:minimal_gas_costs} and \ref{table:minimal_gas_costs_er}
or Figures~\ref{fig:minimal_gas_hist} and \ref{fig:minimal_gas_hist_er}.
\UnrolledThree{} has a plurality of the minimal cost for random values.

\input{tables/minimal_gas_er.tex}
\input{plots/minimal_hist_plots_er.tex}

\subsection{Conclusion}

This extended analysis gives additional insight but with the same result:
\UnrolledThree{} is the best algorithm
for a generic \texttt{uint256} value.
